---
title: "PS5"
author: "Reena"
date: "2023-02-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load and Check Data

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)

data <- read.delim("gapminder.csv.bz2")

nrow(data)

ncol(data)
```
There are 13055 rows, 25 columns

```{r}
rand_rows <- sample(nrow(data), 5)
data[rand_rows, ]
```
It looks fine.


## Descriptive Statistics

```{r}

length(unique(data$iso2))
length(unique(data$name))
length(unique(data$iso3))
```
1) It looks like there are 253 iso3, 249 iso2, and 250 names. 

```{r}
countries <- unique(data[, c("name", "iso2", "iso3")])
```
It looks like some iso3 names has missing iso2 and names. Namibia does not have a corresponding iso2.

```{r}
iso2names <- data %>% 
  group_by(iso2) %>% 
  summarise(num = n_distinct(name))
```
2) a. Looks like there is no iso2 codes that correspond to more than one name. However, there are two blank iso2 lines.

```{r}
iso3names <- data %>% 
  group_by(name) %>% 
  summarise(num = n_distinct(iso3))
```
 b. There are 4 blank lines. Looks like below four countries have alternative names:
1988, Cocos (Keeling) Islands, CCK/CHANISL
3974, Gabon, GAB/GBM
6261, Republic of Korea, KOR/KOS
8666, Netherlands, NLD/NLD CURACAO

```{r}
cleaned_data <- data[complete.cases(data$time),]
max(cleaned_data$time)
min(cleaned_data$time)
```
3) Excluding the NA values, the maximum year is 2019 and the minimum year is 1960.


## CO2 Emissions
```{r}
sum(is.na(data$co2))
sum(is.na(data$co2_PC))
```
1) There are 2658 missing data for co2 and 2661 missing data for co2_PC

```{r}
missing_counts <- data %>%
  group_by(time) %>%
  summarise(co2_missing = sum(is.na(data$co2)),
            co2_pc_missing = sum(is.na(data$co2_PC)))

most_missing_co2 <- missing_counts[which.max(missing_counts$co2_missing), "time"]
most_missing_co2_pc <- missing_counts[which.max(missing_counts$co2_pc_missing), "time"]

most_missing_co2
most_missing_co2_pc
```
I had to regroup the dataset again by missing_counts. I found out that for both co2 and co2_PC columns, 1960 has the most missing data.

```{r}
data %>% 
  filter(iso3 ==c( "USA", "CHN", "IND",  "GBR", "KOR", "SGP")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2, col=iso3))+
  geom_point(aes(x=time, y = co2, col=iso3))

```
2) This is a plot of total CO2 emissions over time for the U.S, China, India, Korea, Great Britain, 
and Singapore. I found that the CO2 emission of USA and China are significantly high no matter
who they compared to.

```{r}
data %>% 
  filter(iso3 ==c( "USA", "CHN", "IND",  "GBR", "KOR", "SGP")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2_PC, col=iso3))+
  geom_point(aes(x=time, y = co2_PC, col=iso3))
```
3) This is a plot for CO2 per capita for these countries. I can see that the US is very high
for CO2 emissions per capita. It does not even intersect with other countries. China's CO2
emissions per capita is actually lower than a lot other countries, but still higher than that
of Inida.

```{r}
ave_co2_pc_region_2016 <- data %>%
  filter(time == 2016) %>%
  group_by(region) %>%
  summarize(ave_co2_pc = mean(co2_PC, na.rm = TRUE))
ave_co2_pc_region_2016$time <- 2016
ave_co2_pc_region_2016

ave_co2_pc_region_1960 <- data %>%
  filter(time == 1960) %>%
  group_by(region) %>%
  summarize(ave_co2_pc = mean(co2_PC, na.rm = TRUE))
ave_co2_pc_region_1960$time <- 1960
ave_co2_pc_region_1960
```
4) Europe has the highest average CO2_PC in 2016, followed by Americas and Asia. 
In 1960, Americas has the highest average CO2_PC, followed by Europe and Oceania.

```{r}
ave_co2_pc_comb <- rbind(ave_co2_pc_region_1960, ave_co2_pc_region_2016)
ggplot(ave_co2_pc_comb, aes(x = region, y = ave_co2_pc, fill = factor(time))) + 
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("purple", "orange")) +
  labs(x = "Region", y = "Average CO2 (per capita)", fill = "Year")

```

