---
title: "PS5"
author: "Reena"
date: "2023-02-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load and Check Data
2) (2pt) Load data. How many rows/columns do we have?
```{r, echo=FALSE}
library(dplyr)
library(ggplot2)

data <- read.delim("gapminder.csv.bz2")

nrow(data)

ncol(data)
```
*There are 13055 rows, 25 columns.*

3) (2pt) Print a small sample of data. Does it look OK?
```{r}
rand_rows <- sample(nrow(data), 5)
data[rand_rows, ]
```
*It looks fine.*


## Descriptive Statistics

1) (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.

```{r}

length(unique(data$iso2))
length(unique(data$name))
length(unique(data$iso3))
```
*It looks like there are 253 iso3, 249 iso2, and 250 names.*

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?

(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
correspond to more than one name? What are these countries?
(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
```{r}
countries <- unique(data[, c("name", "iso2", "iso3")])
```
*It looks like some iso3 names has missing iso2 and names. Namibia does not have a corresponding iso2.*

```{r}
iso2names <- data %>% 
  group_by(iso2) %>% 
  summarise(num = n_distinct(name))
```
*Looks like there is no iso2 codes that correspond to more than one name. However, there are two blank iso2 lines.*

```{r}
iso3names <- data %>% 
  group_by(name) %>% 
  summarise(num = n_distinct(iso3))
```
*There are 4 blank lines. Looks like below four countries have alternative names:*
*1988, Cocos (Keeling) Islands, CCK/CHANISL*
*3974, Gabon, GAB/GBM*
*6261, Republic of Korea, KOR/KOS*
*8666, Netherlands, NLD/NLD CURACAO*


3. (2pt) What is the minimum and maximum year in these data?
```{r}
cleaned_data <- data[complete.cases(data$time),]
max(cleaned_data$time)
min(cleaned_data$time)
```
*Excluding the NA values, the maximum year is 2019 and the minimum year is 1960.*


## CO2 Emissions
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
sum(is.na(data$co2))
sum(is.na(data$co2_PC))
```
*There are 2658 missing data for co2 and 2661 missing data for co2_PC。*

```{r}
missing_counts <- data %>%
  group_by(time) %>%
  summarise(co2_missing = sum(is.na(data$co2)),
            co2_pc_missing = sum(is.na(data$co2_PC)))

most_missing_co2 <- missing_counts[which.max(missing_counts$co2_missing), "time"]
most_missing_co2_pc <- missing_counts[which.max(missing_counts$co2_pc_missing), "time"]

most_missing_co2
most_missing_co2_pc
```
*I had to regroup the dataset again by missing_counts. I found out that for both co2 and co2_PC columns, 1960 has the most missing data.*

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.
```{r}
data %>% 
  filter(iso3 ==c( "USA", "CHN", "IND",  "GBR", "KOR", "SGP")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2, col=iso3))+
  geom_point(aes(x=time, y = co2, col=iso3))

```
*This is a plot of total CO2 emissions over time for the U.S, China, India, Korea, Great Britain,*
*and Singapore. I found that the CO2 emission of USA and China are significantly high no matter*
*who they compared to.*

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
data %>% 
  filter(iso3 ==c( "USA", "CHN", "IND",  "GBR", "KOR", "SGP")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2_PC, col=iso3))+
  geom_point(aes(x=time, y = co2_PC, col=iso3))
```
*This is a plot for CO2 per capita for these countries. I can see that the US is very high *
*for CO2 emissions per capita. It does not even intersect with other countries. China's CO2 *
*emissions per capita is actually lower than a lot other countries, but still higher than that *
*of India.*

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
```{r}
ave_co2_pc_region_2016 <- data %>%
  filter(time == 2016) %>%
  group_by(region) %>%
  summarize(ave_co2_pc = mean(co2_PC, na.rm = TRUE))
ave_co2_pc_region_2016$time <- 2016
ave_co2_pc_region_2016 <- ave_co2_pc_region_2016 %>%
  filter(ave_co2_pc != 19.396490)
ave_co2_pc_region_2016

ave_co2_pc_region_1960 <- data %>%
  filter(time == 1960) %>%
  group_by(region) %>%
  summarize(ave_co2_pc = mean(co2_PC, na.rm = TRUE))
ave_co2_pc_region_1960$time <- 1960
ave_co2_pc_region_1960 <- na.omit(ave_co2_pc_region_1960)
ave_co2_pc_region_1960
```
*Europe has the highest average CO2_PC in 2016, followed by Americas and Asia. *
*In 1960, Americas has the highest average CO2_PC, followed by Europe and Oceania.*


5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r}
ave_co2_pc_comb <- rbind(ave_co2_pc_region_1960, ave_co2_pc_region_2016)
ggplot(ave_co2_pc_comb, aes(x = region, y = ave_co2_pc, fill = factor(time))) + 
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("purple", "orange")) +
  labs(x = "Continent", y = "Average CO2 (per capita)", fill = "Year")

```
*Here is a plot for average CO2 emissions per capita across continents in 1960 and 2016.*


6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).
```{r}
data_filtered <- data %>% filter(!is.na(co2_PC))
data_filtered_summary <- data_filtered %>% filter(time == 2016) %>%
  group_by(region) %>% 
  slice_max(order_by = co2_PC, n = 3) %>% 
  bind_rows(data_filtered %>% filter(time == 2016) %>% 
              group_by(region) %>% 
              slice_min(order_by = co2_PC, n = 3))
```
*The largest 3 CO2 emitters:*
*Africa：South Africa, Libya, Seychelles*
*Americas: Trinidad and Tobago, USA, Canada*
*Asia: Qatar, Kuwait, Bahrain*
*Europe: Gibraltar, Luxembourg, Faroe Islands*
*Oceania: New Caledonia, Australia, Palau*

*The smallest 3 CO2 emitters:*
*Africa: Congo, Somalia, Burundi*
*Americas: Haiti, Nicaragua, Honduras*
*Asia: Afghanistan, Nepal, Yemen*
*Europe: Liechtenstein, Albania, Republic of Moldova.*
*Oceania: Solomon Islands, Vanuatu, Kiribati*

## GDP per capita

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent.
```{r}
data_1960 <- data %>% filter(time == 1960)
ggplot(data_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point(alpha=0.7) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  scale_size_continuous(range = c(1, 10), breaks = c(1000000, 5000000, 10000000, 50000000)) +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  labs(title = "GDP per capita versus life expectancy by country (1960)",
       subtitle = "Point size indicates total population, color indicates continent") +
  theme_bw()

```
*Here is a scatterplot of GDP per capita versus life expectancy by country in 1960. Europe has relatively high* *life expectancy. Asia has a lot of people but lower life expectancy. The Americas are in between Asia and Europe.* 
*The higher the GDP per capita, the higher the life expectancy.*


2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}
data_1960 <- data %>% filter(time == 2019)
ggplot(data_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point(alpha=0.7) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  scale_size_continuous(range = c(1, 10), breaks = c(1000000, 5000000, 10000000, 50000000)) +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  labs(title = "GDP per capita versus life expectancy by country (2016)",
       subtitle = "Point size indicates total population, color indicates continent") +
  theme_bw()
```
*Here is a similar plot for 2016.* 

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

*Africa did not make so much progress. Asia's life expectancy went up, *
*some Asia countries move to the top with Americas and Europe.*
*GDP per capita still seems to correlate with life expectancy. *
*Overall in the past 60 years, GDP per capita went up significantly, but*
*how the GDP is distributed between continents did  not change too much.*
*Africa is still at the bottom, Americas and Europe are still at top. *


4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
```{r}
avg_lifeexp_1960 <- data %>%
  filter(time == 1960) %>% 
  group_by(region) %>% 
  summarize(avg_lifeexp = mean(lifeExpectancy, na.rm = TRUE))
avg_lifeexp_1960

# Compute average life expectancy by continent in 2019
avg_lifeexp_2019 <- data %>%
  filter(time == 2019) %>% 
  group_by(region) %>% 
  summarize(avg_lifeexp = mean(lifeExpectancy, na.rm = TRUE))
avg_lifeexp_2019
```
*Yes, it is consistent with what I see on figures.*

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
```{r}
growth <- data %>%
  group_by(region) %>%
  mutate(growth = (lifeExpectancy - 
  lag(lifeExpectancy))/lag(lifeExpectancy)) %>%
  filter(time > 1960 & time < 2019) %>%
  summarize(avg_growth = mean(growth, na.rm=TRUE)) %>%
  arrange(desc(avg_growth))
growth
```
*The growth is the highest in Africa. Since Europe's LE is already high, there is probably not a *
*lot of capacity for them to grow further.*


6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
GDP_1960_2019 <- data %>%
  filter(time %in% c(1960, 2019))

ggplot(GDP_1960_2019, aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge", alpha = 0.7, bins = 20) +
  scale_fill_manual(values = c("purple", "pink"), name = "Year") +
  labs(title = "GDP per capita 1960 vs 2019", x = 
         "GDP per capita", y = "Frequency")
```
*Here is the histogram of GDP per capita for years of 1960 and 2019.*


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
```{r}
rank_1960 <- data %>%
  filter(time == 1960) %>%
  arrange(desc(lifeExpectancy)) %>%
  mutate(rank = rank(-lifeExpectancy, na.last = "keep")) %>%
  filter(name == "United States of America") 
rank_1960

rank_2019 <- data %>%
  filter(time == 2019) %>%
  arrange(desc(lifeExpectancy)) %>% 
  mutate(rank = rank(-lifeExpectancy, na.last = "keep")) %>% 
  filter(name == "United States of America")
rank_2019
```
*The ranking is 18 in 1960 and 47 in 2019.*


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
```{r}
n_countries_1960 <- data %>%
  filter(time == 1960 & !is.na(lifeExpectancy)) %>% 
  nrow()
rank_1960$rank/n_countries_1960

n_countries_2019 <- data %>%
  filter(time == 2019 & !is.na(lifeExpectancy)) %>% 
  nrow()
rank_2019$rank/n_countries_2019
```
*I got 0.0952381 for 1960 and 0.2373737 for 2019. Slightly different from the hint.*
*Looks like the US is still falling significantly even with the relative rank.*

# It took me 7 hours to do this homework.